\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography

### 2012-12-18 flux.tex
Revise incidence comparison with better numbers for year-to-year drift.
Instead of taking mean location averaged across posterior for year (i-1) and
comparing this to mean location averaged across posterior for year i.  I take
each posterior slice and compare mean location of year (i-1) to year i. I then
average this measurements of year-to-year antigenic drift.

%%% jumps %%%
\begin{figure}[h]
	\centering		
	\includegraphics[width=0.55\textwidth]{figures/jumps}
	\caption{\textbf{Year-to-year antigenic drift of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
	Year-to-year antigenic drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of viruses collected in year $i$ compared to the mean of antigenic dimension 1 of viruses collected during the previous year $i-1$.
	Bars represent interquartile range across MCMC samples.
	} 
	\label{jumps} 
\end{figure}

@@

We investigated the relative rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 1999 to 2008, finding a general correspondence between antigenic drift in dimension 1 and relative incidence within each influenza clade (Figure~\ref{incidence}).
From 1999 to 2008, A/H3N2 accounted for the majority of antigenic drift (46\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 21\%, 17\% and 16\%, respectively.
Similarly, A/H3N2 was responsible for the majority of incidence (54\%) in the USA during this time period.
Influenza A/H1N1 accounted for 23\% of incidence, B/Vic accounted for 12\% of incidence and B/Yam accounted for 11\% of incidence.
These proportions are significantly similar to one another (randomization test, $p = 0.015$).

%%% incidence %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/incidence}
	\caption{\textbf{Relative antigenic drift and relative incidence of A/H3N2, A/H1N1, B/Vic and B/Yam from 1999 to 2008.} 
	(A) Relative antigenic drift of influenza clades.
	Relative antigenic drift for A/H3N2 is measured as the drift of A/H3N2 viruses divided by the sum of drift for A/H3N2, A/H1N1, B/Vic and B/Yam.
	Drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of viruses collected in year $i$ compared to the mean of antigenic dimension 1 of viruses collected during the previous year $i-1$.
	For example, `99' signifies difference in antigenic dimension 1 between 1998 and 1999.
	(B) Relative incidence of influenza clades determined through relative isolation counts in the USA for winter influenza seasons.
	Here, `99' refers to the 1999/2000 influenza season.
	(C) Correlation of relative antigenic drift and relative incidence in each influenza clade.
	Scatterplots represent drift from year $i-1$ to year $i$ compared to incidence in winter season beginning in year $i$, e.g. drift from 1998 to 1999 compared to incidence in the 1999/2000 influenza season.
	} 
	\label{incidence} 
\end{figure}

Additionally, year-to-year comparisons show a similar pattern (Figure~\ref{incidence}). 
Here, we define antigenic drift from year $i-1$ to year $i$ as the difference between the average location in dimension 1 of viruses sampled in year $i-1$ and the average location in dimension 1 of viruses sampled in year $i$.
For instance, we see that 97\% of the antigenic drift between 1998 and 1999 is attributable to A/H3N2 and subsequently the 1999-2000 winter influenza season is dominated by A/H3N2.
Similarly, between 1999 and 2000, most antigenic drift is attributable to A/H1N1 and B/Yam, and the 2000-2001 season is dominated by these same two clades.
Here, we find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.67$ (Pearson correlation, $p = 1.00 \times 10^{-6}$).
Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.63$, $r_\mathrm{H1} = 0.20$, $r_\mathrm{Vic} = 0.74$ and $r_\mathrm{Yam} = 0.58$.

These findings clearly demonstrate the epidemiological importance of antigenic drift and show that competition between influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam is strongly impacted by levels of antigenic drift within each clade.
Furthermore, these findings suggest that the composition of a particular influenza season could be estimated ahead of time by examining the degree of antigenic drift between influenza clades.

@@

We simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
To do this, we incorporate a prior on virus locations according to the evolutionary process
\begin{equation}
	(X_1,\ldots,X_n) \sim \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd),
\end{equation}
where $\tree$ is a phylogeny specifying tree topology and branch lengths and $\diffusionsd$ is the volatility parameter of the Brownian motion,  set to have equal variance in each antigenic dimension and no correlation across dimensions.
Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST, BEAST17}.
The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.
Parameter $\diffusionsd$ is assumed to follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
Metropolis-Hastings proposals scale $\diffusionsd$.
The probability of observing virus locations $p(\mathbf{X}|\tree,\diffusionsd)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
Thus, under the full model, the posterior probability of observing virus and serum locations given immunological data and phylogeny is factored
\begin{equation}
	p(\mathbf{X},\mathbf{Y} | \mathbf{H},\tree) = \frac{ p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd) \; 
	p(\mathbf{X} | \beta, \vartheta) \;
	p(\mathbf{X} | \tree, \diffusionsd) \; 
	p(\mathbf{Y},\mathbf{S},\mathbf{V},\mdssd,\beta,\vartheta,\diffusionsd)}{ p(\mathbf{H}) }.
\end{equation}
Source code implementing these models has been made fully available as part of the software package BEAST \cite{BEAST, BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).

@@

Antigenic, genetic and surveillance data is available as supplemental information \tbc{PREPARE}.

>>>>

%%% jumps %%%
\begin{figure}[h]
	\centering		
	\includegraphics[width=0.5\textwidth]{figures/jumps}
	\caption{\textbf{Year-to-year antigenic drift of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
	Year-to-year antigenic drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of viruses collected in year $i$ compared to the mean of antigenic dimension 1 of viruses collected during the previous year $i-1$.
	Bars represent interquartile range across MCMC samples.
	} 
	\label{jumps} 
\end{figure}

@@

We investigated the relative rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 1999 to 2008, finding a general correspondence between antigenic drift in dimension 1 and relative incidence within each influenza clade (Figure~\ref{incidence}).
To rule out effects caused by the 2009 pandemic, incidence data were examined only until 2008.
From 1999 to 2008, A/H3N2 accounted for the majority of antigenic drift (51\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 22\%, 15\% and 13\%, respectively.
Similarly, A/H3N2 was responsible for the majority of incidence (54\%) in the USA during this time period.
Influenza A/H1N1 accounted for 23\% of incidence, B/Vic accounted for 12\% of incidence and B/Yam accounted for 11\% of incidence.
These proportions are significantly similar to one another (randomization test, $p = 0.001$).

%%% incidence %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/incidence}
	\caption{\textbf{Relative antigenic drift and relative incidence of A/H3N2, A/H1N1, B/Vic and B/Yam from 1999 to 2008.} 
	(A) Relative antigenic drift of influenza clades.
	Relative antigenic drift for A/H3N2 is measured as the drift of A/H3N2 viruses divided by the sum of drift for A/H3N2, A/H1N1, B/Vic and B/Yam.
	Drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of viruses collected in year $i$ compared to the mean of antigenic dimension 1 of viruses collected during the previous year $i-1$.
	For example, `99' signifies difference in antigenic dimension 1 between 1998 and 1999.
	Years with negative absolute drift were assigned a relative antigenic drift of 0.
	(B) Relative incidence of influenza clades determined through relative isolation counts in the USA for winter influenza seasons.
	Here, `99' refers to the 1999/2000 influenza season.
	(C) Correlation of relative antigenic drift and relative incidence in each influenza clade.
	Scatterplots represent drift from year $i-1$ to year $i$ compared to incidence in winter season beginning in year $i$, e.g. drift from 1998 to 1999 compared to incidence in the 1999/2000 influenza season.
	Years lacking antigenic information (Figure~\ref{jumps}) were not included in the correlation.
	} 
	\label{incidence} 
\end{figure}

Additionally, year-to-year comparisons show a similar pattern (Figure~\ref{incidence}). 
Here, we define antigenic drift from year $i-1$ to year $i$ as the difference between the average location in dimension 1 of viruses sampled in year $i-1$ and the average location in dimension 1 of viruses sampled in year $i$.
For instance, we see that 97\% of the antigenic drift between 1998 and 1999 is attributable to A/H3N2 and subsequently the 1999-2000 winter influenza season is dominated by A/H3N2.
Similarly, between 1999 and 2000, most antigenic drift is attributable to A/H1N1 and B/Yam, and the 2000-2001 season is dominated by these same two clades.
Here, we find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.74$ (Pearson correlation, $p = 2.43 \times 10^{-7}$).
Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.61$, $r_\mathrm{H1} = 0.68$, $r_\mathrm{Vic} = 0.23$ and $r_\mathrm{Yam} = 0.56$.

These findings clearly demonstrate the epidemiological importance of antigenic drift and show that incidence of influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam is strongly impacted by levels of antigenic drift within each clade.
Furthermore, these findings suggest that the composition of a particular influenza season could be estimated ahead of time by examining the degree of antigenic drift between influenza clades.
However, these findings alone do not conclusively show interference between clades; more work needs to be done to establish whether antigenic drift within one influenza clades drives down incidence in other clades.

@@

We simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
To do this, we incorporate a prior on virus locations according to the evolutionary process
\begin{equation}
	(X_1,\ldots,X_n) \sim \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd),
\end{equation}
where $\tree$ is a phylogeny specifying tree topology and branch lengths and $\diffusionsd$ is the volatility parameter of the Brownian motion,  set to have equal variance in each antigenic dimension and no correlation across dimensions.
Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST, BEAST17}.
The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.
Parameter $\diffusionsd$ is assumed to follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
Metropolis-Hastings proposals scale $\diffusionsd$.
The probability of observing virus locations $p(\mathbf{X}|\tree,\diffusionsd)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
Thus, under the full model, the posterior probability of observing virus and serum locations given immunological data and phylogeny is factored
\begin{equation}
	p(\mathbf{X},\mathbf{Y} | \mathbf{H},\tree) = \frac{ p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd) \; 
	p(\mathbf{X} | \beta, \vartheta) \;
	p(\mathbf{X} | \tree, \diffusionsd) \; 
	p(\mathbf{Y},\mathbf{S},\mathbf{V},\mdssd,\beta,\vartheta,\diffusionsd)}{ p(\mathbf{H}) }.
\end{equation}

@@

\subsection*{Implemention details}

\tbc{PREPARE}.

\subsection*{Availability}

Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST, BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
Antigenic, genetic and surveillance data is available as supplemental information \tbc{PREPARE}.
Also included as supplemental information are XML control files to load antigenic and phylogenetic data into BEAST and perform cartographic analyses \tbc{PREPARE}.
